plyranges: a grammar for transforming genomics data

### 2018-04-17 paper.Rmd
fleshing out example 2

library(AnnotationHub)
q <- AnnotationHub()
bw_file <- q[["AH33458"]] 
```
# Introduction

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tibble)
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)
```

@@

Here we provide examples on how to use the `plyranges` DSL to construct
genomic data workflows.

@@

We use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
@Roadmap_Epigenomics_Consortium2015-pr. We choose to focus on assays for primary
T CD8+ memory cells from peripheral blood. We can then read the BigWig file 
corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation mark 
over chromosome 10.

@@

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

@@

Returning to the GRanges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing the maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

@@

Finally, the overlap inner join is used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contain
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

@@

(example of group_by_overlaps + summarise)

@@

The _plyranges_ package is available on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
_SummarizedExperiment_ class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it to prepare data for the visualisation of multimodal biological 
datasets.

>>>>

library(AnnotationHub)
q <- AnnotationHub()
# T-cell BW file
bw_file <- q[["AH33458"]] 
```
# Introduction

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tibble)
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)

knitr::kable(feature_tbl, booktabs = TRUE)
```

@@

Here we provide examples on how to use the `plyranges` DSL to construct
genomic data workflows and highlight its interoperability with existing
Bionconductor packages.

@@

The Bioconductor package `AnnotationHub` @R-ahub  can be used to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
@Roadmap_Epigenomics_Consortium2015-pr. Here we focus on assays for primary
T CD8+ memory cells from peripheral blood. Using `plyranges` we will read the 
BigWig file  corresponding to the H3 lysine 27 trimethylation (H3K27Me3) 
methylation mark over chromosome 10.

@@

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
```

@@

Returning to the GRanges object containing normalised coverage scores, 
we filter to find the coordinates of the peak 
containing the maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

@@

Finally, the overlap inner join is used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contain
the max peak on chromosome 10 (figure \ref{fig:peak-viz}).

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved via
the `group_by_overlaps` operator. Continuing with the data from the 
Human Epigenome Roadmap Consortium data, we can count the number of reads
that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
marks from the H1 cell line. We extract reads that have a mapping quality
score greater than 20:

```{r}
h1_bam <- read_bam("h1.bam") %>%
  filter(mapq > 20) 
```

Note that the BAM file is only read into memory once we perform an operation
on it. Next we generate our genomic bins using the `tile` function
from `GenomicRanges`:

```{r}
bins <- tile(h1_bam, width = 10000)
```

Finally, we can use `group_by_overlaps` with `summarise` to compute the
total number of reads with in each window.

```{r}
h1_bam_read_summary <- h1_bam %>%
  group_by_overlaps(bins) %>%
  summarise(n_reads = n())
```

@@

The `plyranges` package is available on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
SummarizedExperiment class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it to prepare data for the visualisation of multimodal biological 
datasets.
